Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-16 thesis/02-cosine.Rmd
Adds fixed chapter 2 and discussion

@@

In the article there was no analysis of individual distances but the general look at the data through the usage of mean. The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race. In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype. They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam). Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.

@@

MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing". High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it somewhat larger. Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.

This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random. It seems like similar practice could be beneficial for metrics aimed at measuring similarity between words.

<!-- TODO: numerical test if this distance is significant? check mean mean cosine distances of some random words? -->

@@

<!-- TODO: -->
<!-- 1. https://www.cs.princeton.edu/courses/archive/fall13/cos521/lecnotes/lec11.pdf -->
<!-- 2. https://stats.stackexchange.com/questions/341535/curse-of-dimensionality-does-cosine-similarity-work-better-and-if-so-why -->
<!-- 3. https://stats.stackexchange.com/questions/21547/distance-metric-and-curse-of-dimensions -->
<!-- + check if it may be proved in the code -->

Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
"https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/"
as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms. One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data. At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to this article "https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out"
there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.

<!-- Add more information on the experiment -->

In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation. In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.

@@

<!-- textbf{Control group}

How to properly prepare control group in terms of quality and quantity?
<!-- + check if it may be proved in the code -->

<!--  \textbf{Verifying other similarity meaures}

Besides cosine similarity, there are other methods used to find the similarity between vectors.

a) TS-SS 
todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering

b) WEAT test 
<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->

<!-- [to check https://github.com/taki0112/Vector_Similarity] -->

<!-- TODO: check in code if they make more sense -->

<!-- ## Bibliography
- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->

<!-- ## *** Additional -->

<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->

<!-- Assumptions: -->
<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->

<!-- TODO: check if it is useful for the analysis of black is .. -->

<!-- ### Identifying the bias subspace  -->

<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->

<!-- Resources to analyze: -->
<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->

<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->

<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->

<!-- Critique of the usage of analogies as a proof for gender biases -->

<!-- TODO: check if there is other argument than that algorithm does not allow it -->

<!-- general thoughts 
https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs

-
-
-
-->

>>>>

One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements. The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side. As shown in [Statexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.

@@

In the article there was no analysis of individual distances but the general look at the data through the usage of mean. The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race. In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype. They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam). Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.

@@

MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it somewhat larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.

@@

In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
[Analyticsindiamag Curse of dimensionality](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms. One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data. At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.

@@

### 2021-06-16 thesis/04-discussion.Rmd
Adds fixed chapter 2 and discussion

# Discussion

@@

- the distributions of cosine distances of whole population and individual words can vary enormously
- including uncertainty shows how in some cases it is not clear why a word is according to this method
classified as anomalous
- debiasing with the method provided by @Manzini2019blackToCriminal does not solve the bias issues

One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal. One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words. However what does it mean to be close to 0? If the averaged cosine similarity is equal to 80, then should we still debiase it? It is unclear what the criteria are. On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set. On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.

We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.

@@

The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects. 
In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.

<!-- 
remarks:
- delete papers from github/hide them
-

@@

>>>>

---
output:
  pdf_document: default
  html_document: default
---
# Discussion

@@

<!-- Looking into specific values -->
As we pointed out, if there is no control of what cosine distance values individual word pairs have, then there is less
understanding of the exact relations between words. If the protected words that we chose do not have high similarity with harmful stereotypes, then one should consider at least two scenarios. The first one is that the choice of protected words and attributes may be corrupted. The second one is that the metric is not able to catch the hidden bias properly. In both cases it is essential to take a look at the individual values before averaging them or aggregating in other ways.

<!-- Problem with gender words --> 
One has to remember also about the specific situation present in gender dataset. There surprisingly one could observe high cosine distances values between some female stereotypical professions and male protected words. This leads to new ideas of how to understand the bias origins and its detection. To start with, cosine distance seems to catch the information from the vectors of words co-occurrence and not the semantic similarity. If word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading. What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations. What is more, this highlights again the need for control group. One could check each stereotypical attribute with each protected word but not in order to average the values. One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.

<!-- debiasing and uncertainty -->
Debiasing with the method provided by @Manzini2019blackToCriminal shows how it is unpredictable in terms of the results. 
In some cases the final cosine distance is indeed close to neutral words, but in other cases the change is minor. What is more, including uncertainty shows how in some cases it is not clear why a word is according to this method
classified as anomalous. The fact that uncertainty for `assosiated` and `different` class is so high leads to the situation where in the boundaries of an uncertainty of some attributes cosine distance, one can find the neutral words as well. It makes the usage and interpretation of this metric extremely complicated and unclear.

<!-- debiasing --> 
Additionally, one cannot be sure if the bias is still preserved after the debiasing. The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed. One could argue that maybe it is not clear how to measure bias presence and removal but the method for debiasing works fine. Notwithstanding, it is showed in researches such as @Gonen2019Lipstick, that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods. Therefore it seems to not be justified to claim that the method works properly without precise methods to verify this.

<!-- no baseline for the interpretation of the results -->
One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal. One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words. However what does it mean to be close to 0? If the averaged cosine similarity is equal to 80, then should we still debiase it? It is unclear what the criteria are. On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set. On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.

<!-- Baysian method --> 
We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.

// see the chapter 3 to write more

@@

The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.

In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.

<!-- 
remarks:
- delete papers from github/hide them
- should we quote "debiasing" or without quoting

@@

<!-- textbf{Control group}

How to properly prepare control group in terms of quality and quantity?
<!-- + check if it may be proved in the code -->

<!--  \textbf{Verifying other similarity meaures}

Besides cosine similarity, there are other methods used to find the similarity between vectors.

a) TS-SS 
todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering

b) WEAT test 
<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->

<!-- [to check https://github.com/taki0112/Vector_Similarity] -->

<!-- TODO: check in code if they make more sense -->

<!-- ## Bibliography
- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->

<!-- ## *** Additional -->

<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->

<!-- Assumptions: -->
<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->

<!-- TODO: check if it is useful for the analysis of black is .. -->

<!-- ### Identifying the bias subspace  -->

<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->

<!-- Resources to analyze: -->
<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->

<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->

<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->

<!-- Critique of the usage of analogies as a proof for gender biases -->

<!-- TODO: check if there is other argument than that algorithm does not allow it -->

<!-- general thoughts 
https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs

-
-
-
-->

### 2021-06-16 thesis/_book/_main.tex
Adds fixed chapter 2 and discussion

@@

It is worth diving deeper into some observations that arise while analyzing the results.

\begin{itemize}
\tightlist
\item
  the distributions of cosine distances of whole population and individual words can vary enormously
\item
  including uncertainty shows how in some cases it is not clear why a word is according to this method
  classified as anomalous
\item
  debiasing with the method provided by \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) does not solve the bias issues
\end{itemize}

One should remember that there is no baseline for the interpretation of the averaged cosine distances in (\protect\hyperlink{ref-Manzini2019blackToCriminalToCriminal}{\textbf{Manzini2019blackToCriminalToCriminal?}}). One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words. However what does it mean to be close to 0? If the averaged cosine similarity is equal to 80, then should we still debiase it? It is unclear what the criteria are. On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set. On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.

\hypertarget{what-we-propose-and-how-it-can-help}{%
\section{What we propose and how it can help:}\label{what-we-propose-and-how-it-can-help}}

\begin{itemize}
\tightlist
\item
\end{itemize}

\hypertarget{summary}{%
\chapter{Summary}\label{summary}}

\hypertarget{future-work}{%
\section{Future work:}\label{future-work}}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@
